An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-23 scagnostics.Rmd
References and comments added.

@@

@@

  - smoothness: the index function should evolve smoothly between nearby projections, i.e. on an interpolated path we expect to observe slow and consistent change in the index function. On the other hand if two similar projections produce very different index values, i.e. we observe sharp jumps when looking at the evolution of the index function, and in particular if those fluctuations dominate the overall evolution, the index is not suitable for projection pursuit.
  - speed: the computation time is an important factor in particular when the guided tour is used to view the optimisation in real time. Note that longer computation times may be acceptable when first recording the optimisation and replaying the guided tour once the maximum has been identified.
  - ability to find fine and broad structure: We consider a broad structure one that is roughly observable over a large angle around the optimal projection, it should in general be easy to identify and a good ppi should show smooth increase over a large number of projections as we are approaching the optimal view. On the other hand fine structures are difficult to identify as they are only visible close to the optimal view. This poses problems in particular for the optimization, but we may use index definitions to amplify larger index values while suppressing smaller ones, such that the optimization will be less sensitive to noise.
  - distribution of index values: the distribution of index values when studying data points without structure should have a low variance, and be clearly separated from index values obtained when looking at structured data distributions.
  - rotation invariance: in the tour there is an assumed degeneracy between all projections obtained when rotating within the projected space only, for 2-d projections it means that we cannot assing x- and y- axis without ambiguity. A projection pursuit index should therefore be invariant under rotation of the basis in the projection plane. By costruction this is not the case for most index functions defined above. (One way around could be profiling the function over all possible rotations in the plane, howevere computing time may be a major concern in this case.) (Other examples where rotation invariance is important include image recognition and texture classification, rotation angle may be built in parameter for summary calculation, e.g. https://doi.org/10.1016/S0146-664X(75)80008-6, other example is interpolating over angles when building NN, see e.g. https://arxiv.org/pdf/1805.12301.pdf)

##Simulation study
To test the behaviour of the index functions we use three types of simulated data sets based on different underlying distributions, and with different built-in features. They are defined in $n=6$ dimensions with $p$ randomly selected points $x_i, i = 1,..,n$ according to:

>>>>

\textcolor{red}{\cite{jones87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall intervall is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardising the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardisation procedure, it would most likely be unwelcome effects?}

@@

\textcolor{red}{\cite{posse95b} discusses the optimisation, in particular that for most index functions and optimisers results are too local, largely dependent on starting point. Fiedmans suggestion of hybrid steppind search + steepest-ascent still remains fairly local. \cite{posse90} suggested random search instead, which seems to perform better, but did not find pdf of that article, so not so sure what he actually does.}

@@

\textcolor{red}{\cite{posse95b} lists five criteria: minimized by normal distribution, approximately affine invariant, consistent, resistent to features in the tail, fast computation. Comment that different index values for the same plane will make optimisation less efficient. \cite{jones87} moreover lists rotation and translation invariance (independence of choice of basis in the plane), and also mentions conditions for numeric optimisation, i.e. continuous and depending on optimisation methods also differentiable. Interesting: according to \cite{jones87} the original index proposed by Friedman and Tukey did not have property of rotation invariance.}

  - smoothness: the index function should evolve smoothly between nearby projections, i.e. on an interpolated path we expect to observe slow and consistent change in the index function. On the other hand if two similar projections produce very different index values, i.e. we observe sharp jumps when looking at the evolution of the index function, and in particular if those fluctuations dominate the overall evolution, the index is not suitable for projection pursuit.
  - speed: the computation time is an important factor in particular when the guided tour is used to view the optimisation in real time. Note that longer computation times may be acceptable when first recording the optimisation and replaying the guided tour once the maximum has been identified.
  - ability to find fine and broad structure: We consider a broad structure one that is roughly observable over a large angle around the optimal projection, it should in general be easy to identify and a good ppi should show smooth increase over a large number of projections as we are approaching the optimal view. On the other hand fine structures are difficult to identify as they are only visible close to the optimal view. This poses problems in particular for the optimization, but we may use index definitions to amplify larger index values while suppressing smaller ones, such that the optimization will be less sensitive to noise. \textcolor{red}{there are two different considerations, one is that described above, it relates to the concept of squint angle discussed in \cite{posse95b}. On the other hand should also consider detectin of "fine" structure vs noise, this would be controlled e.g. by the allowed number of bins for the MINE index functions, additionally this will impact computing times, i.e. trade accuracy for efficiency}
  - distribution of index values: the distribution of index values when studying data points without structure should have a low variance, and be clearly separated from index values obtained when looking at structured data distributions.
  - rotation invariance: in the tour there is an assumed degeneracy between all projections obtained when rotating within the projected space only, for 2-d projections it means that we cannot assing x- and y- axis without ambiguity. A projection pursuit index should therefore be invariant under rotation of the basis in the projection plane. By costruction this is not the case for most index functions defined above. (One way around could be profiling the function over all possible rotations in the plane, howevere computing time may be a major concern in this case.) (Other examples where rotation invariance is important include image recognition and texture classification, rotation angle may be built in parameter for summary calculation, e.g. https://doi.org/10.1016/S0146-664X(75)80008-6, other example is interpolating over angles when building NN, see e.g. https://arxiv.org/pdf/1805.12301.pdf)

##Simulation study

\textcolor{red}{\cite{posse95a} uses simulations with "strongly hidden structure", i.e. truncated Gaussian distributions with approximately identity covariance matrix, such that they dont appear in any of the principal component planes.}

To test the behaviour of the index functions we use three types of simulated data sets based on different underlying distributions, and with different built-in features. They are defined in $n=6$ dimensions with $p$ randomly selected points $x_i, i = 1,..,n$ according to:
